---
title: "Intro_to_statitical_learning_with_R_c2_exercises"
output: html_notebook
---

# Conceptual
## Question 1 
Indicate the general expected performance of a flexible statistical learning method whether to be better or worse than an inflexible method. 

a. The sample size **n** is extremely large, and the number of predictors **p** is small. 

A flexible statistical learning method would perform better.They do not assume a shape of the functional relationship between response and predictors.

b. The number of predictors p is extremely large, and the number of observations n is small.

An inflexible method would perform better because flexible methods require a large number of observations.

c. The relationship between the predictors and response is highly
non-linear.

A flexible method would perform better - inflexible methods are restricted by the functional form of f to capture the relationship between predictors and response. A flexible method not restricted by form and tries to model the form based on the data points.

d. The variance of the error terms, i.e. σ2 = Var(ϵ), is extremely
high.

## Question  2
Explain whether each scenario is a classification or regression problem, and indicate whether we are most interested in inference or prediction. Finally, provide n and p

a. **We collect a set of data on the top 500 firms in the US. For each firm we record profit, number of employees, industry and the CEO salary. We are interested in understanding which factors affect CEO salary.**


This is a regression problem as we are trying to infer the relationship with a continuous quantitative response value. *n* is 500. *p* is 3

b. **We are considering launching a new product and wish to know whether it will be a success or a failure. We collect data on 20 similar products that were previously launched. For each product we have recorded whether it was a success or failure, price charged for the product, marketing budget, competition price,and ten other variables.**

This is a classification problem where the response is a binary qualitative response - success or failure. *n* is 20. *p* is 13.

c. **We are interested in predicting the % change in the USD/Euro exchange rate in relation to the weekly changes in the world stock markets. Hence we collect weekly data for all of 2012. For each week we record the % change in the USD/Euro, the % change in the US market, the % change in the British market,and the % change in the German market.**

This is a regression problem - quantitative continuous response variable. *n* is 52. p is *3*.


## Question 3

a. Provide sketch of bias, variance, training error, test error, and Bayes error curves on a single plot. x-axis is flexibility and y-axus represents values of each curve.

b. Explain why each of the five curves has the shape displayed in part a.
- bias: this represents the error approximating a real-life problem which can be complex with a simple model. The less flexible the model the more bias introduced into the problem. Therefore bias will tend to decrease as we increase model flexibility because more flexible models attempt to capture/represent the complexity in the relationships between response and predictors. 
- variance: the variance will tend to increase as the flexibility increases. More flexible methods are sensitive to change in the observations that they model.


```{r}
library(ISLR)
library(ggplot2)
library(dplyr)
#library(plotly)
```

Read in the `College` data set with the `read.csv()` function.
 
```{r}
# read in the College data set
college <- read.csv("../data/College.csv")
```

```{r}
# give each row name corresponding to the appropriate college name
rownames(college) <- college[,1]
```
```{r}
# remove the first column where college names stored
college <- college[,-1]
```

```{r}
# use summary function to produce numerical summary of the variables in the data set
summary(college)
```


- Use the `pairs()` function to produce a scatter plot matrix of the first ten columns or variables of the data.    

```{r}
# create a scatter plot matrix with the first 10 predictors in college datasec
pairs(college[, 2:10])
```
- Use the `plot()` function to produce side by side box plots of `Outstate` vs `Private`.
```{r}
# create multiple plots - 4 rows and 4 columns with margin - top right bottom left
par(mfrow=c(4,4), mar=c(2,2,1,0))

for (i in 2:17)
  boxplot(college[, i] ~ college[, 1], xlab="", main=colnames(college)[i])
```
- Create a new qualitative variable - `Elite`, by *binning* the top10perv variable.

```{r}
# create replicated vector of No with length of college
Elite <- rep("No", nrow(college))
# change elite value to yes if Top10Perc above 50
Elite[college$Top10perc > 50] <- "Yes"
# change Elite to a set of distinct qualitative values - factor
Elite <- as.factor(Elite)
# add Elite to College as a categorical variable
college <- data.frame(college, Elite)
summary(college$Elite)
```
Based on summary of `Elite` column, there are **78** elite universities in the `College` data set. 

Use the 'plot()` function to create side-by-side boxplots of `Outstate` versus `Elite`

```{r}
# create multiple plots - 4 rows and 4 columns with margin - top right bottom left

plot(college$Outstate~ college$Elite)
```


```{r}
par(mfrow=c(2,2))
hist(college$Apps, breaks = 5)
hist(college$Apps, breaks = 10)
hist(college$Apps, breaks = 20)
hist(college$Apps, breaks = 40)
```

```{r}
par(mfrow=c(2,2))
hist(college$PhD, breaks = 5)
hist(college$PhD, breaks = 10)
hist(college$PhD, breaks = 20)
hist(college$PhD, breaks = 40)
```


Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
